Supplementary material for ``Screening Sinkhorn Algorithm for Regularized Optimal Transport''

### 2019-09-24 final_NeurIPS/main.tex
camera ready version NeurIPS

% ready for submission
\usepackage{neurips_2019}

@@

% to compile a camera-ready version, add the [final] option, e.g.:
% \usepackage[]{neurips_2019}

@@

%% algorithms
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{enumitem}

@@

%% math operators
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}
%--------------------------------------------------------------------------------------------------------------------------------%

@@

\author{%
Mokhtar Z. Alaya \\
LITIS EA4108\\
University of Rouen\\
\texttt{mokhtarzahdi.alaya@gmail.com} 
\And
Maxime Bérar\\
LITIS EA4108\\
University of Rouen\\
\texttt{maxime.berar@univ-rouen.fr} \\
\And
Gilles Gasso \\
LITIS EA4108\\
INSA, University of Rouen\\
\texttt{gilles.gasso@insa-rouen.fr} 
\And
Alain Rakotomamonjy\\
LITIS EA4108 \\
University of Rouen\\
and Criteo AI Lab, Criteo Paris \\
\texttt{alain.rakoto@insa-rouen.fr} \\
}

@@

\begin{abstract}
We introduce in this paper a novel strategy for efficiently approximate the Sinkhorn distance between two discrete measures. After identifying neglectible components
of the dual solution of the regularized Sinkhorn problem, we propose to screen those components by directly setting them at that value before entering the Sinkhorn problem. This allows us to solve a smaller Sinkhorn problem while ensuring approximation with provable guarantees.
More formally, the approach is based on a reformulation of \emph{dual of Sinkhorn divergence problem} and on the KKT optimality conditions of this problem, which enable identification of dual components to be screened.
This new analysis leads to the \emph{Screenkhorn} algorithm.
We illustrate the efficiency of Screenkhorn on complex tasks such as dimensionality reduction or domain adaptation involving regularized optimal transport.
\end{abstract}

@@

\section{Conclusion}
The paper introduces a novel efficient approximation of the Sinkhorn divergence
based on a screening strategy. Screening some of the Sinkhorn dual variables
has been made possible by defining a novel constrained dual problem and by 
carefully analyzing its optimality conditions. From the latter, we derived some
sufficient conditions depending on the ground cost matrix, that some dual variables are smaller than a given threshold. Hence, we need just to solve a restricted
dual Sinkhorn problem using an off-the-shelf L-BFGS-B algorithm. We also provide
some theoretical guarantees of the quality of the approximation with respect to
the number of variables that have been screened. Numerical experiments show 
the behaviour of our \emph{Screenkhorn} algorithm and computational time gain it can
achieve when integrated in some complex machine learning pipelines.

% \subsubsection*{Acknowledgments}

% Use unnumbered third level headings for the acknowledgments. All acknowledgments go at the end of the paper. Do not include acknowledgments in the anonymized submission, only in the final paper.

>>>>

% ready for submission
% \usepackage{neurips_2019}

@@

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final]{neurips_2019}

@@

%% algorithms
\usepackage[ruled,linesnumbered,vlined,scleft]{algorithm2e} 
\usepackage{adjustbox}

\newsavebox{\tempbox}

\makeatletter
\let\oldnl\nl% Store \nl in \oldnl
\newcommand{\nonl}{\renewcommand{\nl}{\let\nl\oldnl}}% Remove line number for one line
\makeatother

\let\oldalgorithm\algorithm
\let\oldendalgorithm\endalgorithm

\newcommand\SetAlgoFig{
  \let\algorithm\figure
  \let\endalgorithm\endfigure
}

\newcommand\SetAlgoNoFig{
  \let\algorithm\oldalgorithm
  \let\endalgorithm\oldendalgorithm
}
%%

\usepackage{enumitem}

@@

%% math operators
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}

% small pmatrix
% \usepackage{amsmath}
% \newenvironment{psmallmatrix}
%   {\left(\begin{smallmatrix}}
%   {\end{smallmatrix}\right)}
%--------------------------------------------------------------------------------------------------------------------------------%

@@

\author{%
Mokhtar Z. Alaya \\
LITIS EA4108\\
University of Rouen Normandy\\
\texttt{mokhtarzahdi.alaya@gmail.com} 
\And
Maxime Bérar\\
LITIS EA4108\\
University of Rouen Normandy\\
\texttt{maxime.berar@univ-rouen.fr} \\
\And
Gilles Gasso \\
LITIS EA4108\\
INSA, University of Rouen Normandy\\
\texttt{gilles.gasso@insa-rouen.fr} 
\And
Alain Rakotomamonjy\\
LITIS EA4108 \\
University of Rouen Normandy\\
and Criteo AI Lab, Criteo Paris \\
\texttt{alain.rakoto@insa-rouen.fr} \\
}

@@

\begin{abstract}
We introduce in this paper a novel strategy for efficiently approximating the Sinkhorn distance between two discrete measures. After identifying neglectable components
of the dual solution of the regularized Sinkhorn problem, we propose to screen those components by directly setting them at that value before entering the Sinkhorn problem. This allows us to solve a smaller Sinkhorn problem while ensuring approximation with provable guarantees.
More formally, the approach is based on a new formulation of \emph{dual of Sinkhorn divergence problem} and on the KKT optimality conditions of this problem, which enable identification of dual components to be screened.
This new analysis leads to the \textsc{Screenkhorn} algorithm.
We illustrate the efficiency of \textsc{Screenkhorn} on complex tasks such as dimensionality reduction and domain adaptation involving regularized optimal transport.
\end{abstract}

@@

\section{Conclusion}
The paper introduces a novel efficient approximation of the Sinkhorn divergence
based on a screening strategy. Screening some of the Sinkhorn dual variables
has been made possible by defining a novel constrained dual problem and by 
carefully analyzing its optimality conditions. From the latter, we derived some
sufficient conditions depending on the ground cost matrix, that some dual variables are smaller than a given threshold. Hence, we need just to solve a restricted
dual Sinkhorn problem using an off-the-shelf L-BFGS-B algorithm. We also provide
some theoretical guarantees of the quality of the approximation with respect to
the number of variables that have been screened. Numerical experiments show 
the behaviour of our \textsc{Screenkhorn} algorithm and computational time gain it can
achieve when integrated in some complex machine learning pipelines.

\subsubsection*{Acknowledgments}

This work was supported by grants from the Normandie Projet GRR-DAISI, European funding FEDER DAISI and OATMIL ANR-17-CE23-0012 Project of the French National Research Agency (ANR).

### 2019-09-24 final_NeurIPS/section_2_setup_OT.tex
camera ready version NeurIPS

We briefly present in this section the setup of OT between two discrete measures. We then consider the case when those distributions are only available through a finite number of samples, that is $\mu = \sum_{i=1}^n \mu_i \delta_{x_i} \in \Sigma_n$ and $\nu = \sum_{j=1}^m \nu_i \delta_{y_j} \in \Sigma_m$, where $\Sigma_n$ is the probability simplex with $n$ bins, namely the set of probability vectors in $\R_+^n$, i.e., $\Sigma_n = \{w \in \R_+^n: \sum_{i=1}^n w_i = 1\}.$
We denote their probabilistic couplings set as $\Pi(\mu, \nu) = \{P \in \R_+^{n\times m}, P\mathbf{1}_m = \mu, P^\top \mathbf{1}_n = \nu\}.$ 
\paragraph{Sinkhorn divergence.}

@@

Below we provide the derivation of the dual problem for the regularized OT problem~\eqref{sinkhorn-primal}. Towards this end, we begin with writing its Lagrangian dual function:
\begin{equation*}
  \mathscr{L}(P,y, z) = \inr{C,P} + \eta \inr{\log P, P} + \inr{y, P\mathbf{1}_m - \mu} + \inr{z,P^\top \mathbf{1}_n - \nu}.
\end{equation*}
The dual of Sinkhorn divergence can be derived by solving $\min_{P \in \R_+^{n\times m}}\mathscr{L}(P,y, z)$. It is easy to check that objective function $P\mapsto \mathscr{L}(P,y, z)$ is strongly convex and differentiable. Hence, one can solve the latter minimum by setting $\nabla_P \mathscr{L}(P,y, z)$ to $\mathbf{0}_{n\times m}$. Therefore, we get 
$
  P^\star_{ij} = \exp\Big(- \frac{1}{\eta} (y_i + z_j + C_{ij}) - 1\Big), 
$
for all $i=1, \ldots, n$ and $j=1, \ldots, m$. Plugging this solution,  and setting the change of variables $u = -y/\eta - 1/2$ and $v = - z/\eta - 1/2$, the dual problem is given by
\begin{equation}
\label{sinkhorn-dual}
\min_{u \in \R^n, v\in\R^m}\big\{\Psi(u,v):= \mathbf{1}_n^\top B(u,v)\mathbf{1}_m - \inr{u, \mu} - \inr{v, \nu} \big\},
\end{equation}
where $B(u,v) := \Delta(e^{u}) K \Delta(e^{v})$ and $K := e^{-C/\eta}$ stands for the Gibbs kernel associated to the cost matrix $C$. 
We refer to problem~\eqref{sinkhorn-dual} as the \emph{dual of Sinkhorn divergence}, and its optimal solution $P^\star$ takes the form $P^\star = \Delta(e^{u^\star}) K \Delta(e^{v^\star})$
where the couple $(u^\star, v^\star)$ satisfies:
\begin{align*}
\label{sinkhorn-dual}
  (u^\star, v^\star) &= \argmin_{u \in \R^{n}, v\in \R^m} \{\Psi(u,v)\}.
\end{align*}
Note that the matrices $\Delta(e^{u^\star})$ and $\Delta(e^{v^\star})$ are unique up to a constant factor~\citep{sinkhorn1967}. Moreover, $P^\star$ can be solved efficiently by iterative Bregman projections~\citep{benamou2015IterativeBP} referred as Sinkhorn iterations, and the method is referred as Sinkhorn algorithm which, recently, is proven to achieve a near-$\bigO(n^2)$ complexity~\citep{altschulernips17}.

>>>>

We briefly expose in this section the setup of OT between two discrete measures. We then consider the case when those distributions are only available through a finite number of samples, that is $\mu = \sum_{i=1}^n \mu_i \delta_{x_i} \in \Sigma_n$ and $\nu = \sum_{j=1}^m \nu_i \delta_{y_j} \in \Sigma_m$, where $\Sigma_n$ is the probability simplex with $n$ bins, namely the set of probability vectors in $\R_+^n$, i.e., $\Sigma_n = \{w \in \R_+^n: \sum_{i=1}^n w_i = 1\}.$
We denote their probabilistic couplings set as $\Pi(\mu, \nu) = \{P \in \R_+^{n\times m}, P\mathbf{1}_m = \mu, P^\top \mathbf{1}_n = \nu\}.$ 
\paragraph{Sinkhorn divergence.}

@@

Below we provide the derivation of the dual problem for the regularized OT problem~\eqref{sinkhorn-primal}. Towards this end, we begin with writing its Lagrangian dual function:
\begin{equation*}
  \mathscr{L}(P,w, z) = \inr{C,P} + \eta \inr{\log P, P} + \inr{w, P\mathbf{1}_m - \mu} + \inr{z,P^\top \mathbf{1}_n - \nu}.
\end{equation*}
The dual of Sinkhorn divergence can be derived by solving $\min_{P \in \R_+^{n\times m}}\mathscr{L}(P,w, z)$. It is easy to check that objective function $P\mapsto \mathscr{L}(P,w, z)$ is strongly convex and differentiable. Hence, one can solve the latter minimum by setting $\nabla_P \mathscr{L}(P,w, z)$ to $\mathbf{0}_{n\times m}$. Therefore, we get $  P^\star_{ij} = \exp\big(- \frac{1}{\eta} (w_i + z_j + C_{ij}) - 1\big), 
$ for all $i=1, \ldots, n$ and $j=1, \ldots, m$. Plugging this solution,  and setting the change of variables $u = -w/\eta - 1/2$ and $v = - z/\eta - 1/2$, the dual problem is given by
\begin{equation}
\label{sinkhorn-dual}
\min_{u \in \R^n, v\in\R^m}\big\{\Psi(u,v):= \mathbf{1}_n^\top B(u,v)\mathbf{1}_m - \inr{u, \mu} - \inr{v, \nu} \big\},
\end{equation}
where $B(u,v) := \Delta(e^{u}) K \Delta(e^{v})$ and $K := e^{-C/\eta}$ stands for the Gibbs kernel associated to the cost matrix $C$. 
We refer to problem~\eqref{sinkhorn-dual} as the \emph{dual of Sinkhorn divergence}. Then, the optimal solution $P^\star$ of the primal problem~\eqref{sinkhorn-primal} takes the form $P^\star = \Delta(e^{u^\star}) K \Delta(e^{v^\star})$
where the couple $(u^\star, v^\star)$ satisfies:
\begin{align*}
\label{sinkhorn-dual}
  (u^\star, v^\star) &= \argmin_{u \in \R^{n}, v\in \R^m} \{\Psi(u,v)\}.
\end{align*}
Note that the matrices $\Delta(e^{u^\star})$ and $\Delta(e^{v^\star})$ are unique up to a constant factor~\citep{sinkhorn1967}. Moreover, $P^\star$ can be solved efficiently by iterative Bregman projections~\citep{benamou2015IterativeBP} referred to as Sinkhorn iterations, and the method is referred to as \textsc{Sinkhorn} algorithm which, recently, has been proven to achieve a near-$\bigO(n^2)$ complexity~\citep{altschulernips17}.
